Forecasting the 2024 U.S. Presidential Election: An Analysis for Battleground States\thanks{Code and data are available at: \url{https://github.com/krishnak30/US_elections}.}

### 2024-11-01 paper/paper.tex
commit intermediate files

\begin{document}
\maketitle
\begin{abstract}
This study employs a polls-of-polls approach to predict support for
Kamala Harris and Donald Trump in the 2024 U.S. Presidential Election,
with a focus on key battleground states to forecast the likely winner of
the electoral college. By aggregating multiple polls and applying a
weighted linear regression model with predictors like pollster
reliability, sample size, state, and recency, we estimate a higher
national support for Trump relative to Harris. Our analysis also reveals
a close competition in battleground states, with Trump holding a narrow
lead in Arizona, Georgia, and North Carolina, while Harris leads in
Michigan, Nevada, Pennsylvania, and Wisconsin. These findings underscore
the value of aggregating polls over relying on individual surveys,
offering a more comprehensive and robust forecast of electoral outcomes.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, boxrule=0pt, sharp corners, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

This paper is organized as follows: In \ref{sec-data}, we clean the
data, explore summary statistics, plot distributions of key variables,
and examine relationships between variables. In \ref{sec-model}, we
discuss our forecasting approach, model selection, justification for the
chosen model, and the mechanism of deriving poll weights based on
pollster reliability, ultimately presenting our predictions. Finally, in
\ref{sec-discussion}, we address the broader implications of our
findings, acknowledge limitations, and suggest directions for future
work. \vspace{0.7cm} \# Data \{\#sec-data\} \vspace{0.7cm} \#\#
Measurement

In the dataset of our analysis, the process of measurement begins by
capturing real-world public opinion through surveys. Polling
organizations transform this observed phenomenon into structured data by
recording support percentages, pollster details, and methodological
choices, thus turning public sentiment into quantifiable entries in the
dataset.

Public opinion polls are essential for understanding voter preferences
and electoral dynamics, providing data that informs policy decisions and
campaign strategies. Our data set captures public support for candidates
at specific times, along with information on methodologies, sample
sizes, and pollster ratings.

Polling organizations use various methods, such as App panels, IVR
(Interactive Voice Response), Live phone calls, Text-to-web systems etc.
These methods can influence the validity and reliability of results,
with larger sample sizes generally yielding more reliable estimates.
Some methods, like live phone interviews, may be seen as more
trustworthy than others.

The dataset also includes important measures of pollster performance,
such as numeric grades, poll scores, and transparency scores. These
metrics are calculated by considering bias, race difficulty, predictive
error, and transparency, allowing us to assess the accuracy of each
poll. The ``pollscore,'' for instance, reflects a pollster's past
performance and reliability, adjusted for potential biases.

@@

The raw data for this project, sourced from FiveThirtyEight,
(FiveThirtyEight 2024) underwent a series of cleaning steps to prepare
it for analysis. Initially, duplicate rows were removed to ensure that
only unique observations remained, facilitated by the \texttt{janitor}
package (Firke 2023). A new binary variable, `national', was created to
indicate whether a poll was conducted at the national or state level.
Missing values in the `state' column were replaced with ``Not
Applicable,'' and numeric grades were evaluated to filter out
low-quality pollsters, keeping only those with a numeric grade above 1.
This cutoff was selected to retain mid to high-level pollsters for more
reliable results. These steps were performed using functions from the
\texttt{Tidyverse} package (Wickham et al. 2019). Furthermore, dates
were standardized and converted into a proper format for analysis using
the \texttt{lubridate} package (Grolemund and Wickham 2011). Polls
related to Kamala Harris were retained for further analysis, and
percentage support values were transformed into actual numbers of
supporters based on sample size. Additionally, pollster counts below
five were excluded to focus on more reliable data sources. Polls
regarding Kamala Harris were filtered to include only those conducted
after her official candidacy announcement on July 21, 2024, ensuring the
data reflects post-announcement public sentiment.The cleaned dataset was
saved in Parquet format for efficient storage and retrieval, using the
\texttt{arrow} package (Richardson et al. 2024).

@@

\hypertarget{summary-statistics-of-key-variables}{%
\subsubsection{Summary Statistics of Key
Variables}\label{summary-statistics-of-key-variables}}

@@

For the purposes of analysis in this paper, only a subset of variables
was selected. Two additional variable, national and state, was created
using the existing `state' and `end date' variable. A short description
of each variable of interest is given below.

\begin{itemize}
\item
  \emph{Pollster}: The name of the polling organization conducting the
  poll (e.g., YouGov, RMG Research). This variable helps adjust for
  poll-specific biases.
\item
  \emph{Numeric Grade}: A numeric rating given to the pollster,
  representing the quality or reliability of the organization (e.g.,
  3.0), with higher grades indicating more reliable pollsters.
\item
  \emph{Pollscore}: A score that reflects the reliability of each
  pollster, capturing their historical accuracy and any systematic bias.
  Negative values indicate better predictive accuracy.
\item
  \emph{State}: The U.S. state where the poll was conducted, allowing
  for analysis of regional differences in candidate support.
\item
  \emph{National}: A binary variable indicating whether the poll is
  national (1 for national polls, 0 for state polls).
\item
  \emph{End Date}: The date the poll ended, reflecting the currency of
  the data.
\item
  \emph{Sample Size}: The total number of respondents in the poll.
\item
  \emph{Candidate Name}: The name of the candidate being polled (e.g.,
  Kamala Harris or Donald Trump), identifying the focus of each poll
  result.
\item
  \emph{Pct (Percentage)}: The percentage of respondents in the poll who
  support the specified candidate.
\item
  \emph{Recency}: This variable measures how recent each poll is. It was
  calculated by subtracting the end date of the poll from the end date
  of the most recent poll.
\end{itemize}

@@

From Table~\ref{tbl-sumstat} we see that the average poll in our dataset
has a numeric grade of 2.2, suggesting that most polls are of moderate
to high quality in terms of reliability. The mean pollscore of -0.5
indicates a general trend toward negative values, which suggests these
polls are relatively accurate, as more negative values imply reduced
bias. Additionally, the large average sample size of 1908 respondents
across polls provides a stable basis for our model, reducing variability
and enhancing the reliability of predictions for candidate support.

@@

Figure~\ref{fig-ss} reveals a clear right-skewed distribution. Most of
the sample sizes are clustered between 0 and 3000, with a sharp peak
around 1000-1500. This indicates that the majority of polls are
conducted with smaller sample sizes. As sample size increases, the
frequency significantly drops, with very few polls conducted with sample
sizes larger than 5000, though there are a few outliers with sizes
approaching 10,000 or more. This wide range in sample sizes might affect
the precision of estimates in different polls.

@@

Figure~\ref{fig-reli} displays the distribution of Numeric Grade (left)
and Pollscore (right) across polling organizations. The Numeric Grade
distribution shows that most pollsters are rated between 1.5 and 3, with
peaks around 2.0 and 2.5, indicating a concentration of pollsters with
moderate to high reliability scores. In contrast, the Pollscore
distribution, where lower values indicate higher reliability, shows a
range primarily between -1.5 and 0, with a notable peak around -0.5.
This suggests that while many polls demonstrate relatively low bias,
there is still variability in reliability across organizations. The
distinction between these two metrics emphasizes the need to consider
both quality (Numeric Grade) and potential systematic bias (Pollscore)
when weighting polls in the model.

@@

Figure~\ref{fig-pie} charts illustrate the distribution of polls between
candidates and poll types. The left chart shows that the majority of
polls are conducted at the state level, with a smaller portion being
national. The right chart shows that polling is almost evenly split
between Kamala Harris and Donald Trump. This distribution underscores
the model's balanced approach to capturing state-level nuances as well
as broader national trends, providing a comprehensive view of candidate
support across different contexts

@@

\begin{itemize}
\item
  \(y_i\)\hspace{0pt} is the percentage of support for candidate in poll
  iii,
\item
  \(α\) is the intercept,
\item
  \(β_1\) captures the effect of the polling organization,
\item
  \(β_2​\) captures the effect of the sample size,
\item
  \(β_3\)\hspace{0pt} captures the effect of recency (how recent the
  poll is),
\item
  \(β_4\) capture the effects of the different states
\item
  \hspace{0pt} \(\epsilon_i\) represents the error term, assumed to
  follow a normal distribution with mean 0.
\end{itemize}

@@

\begin{itemize}
\item
  \(w_i\)\hspace{0pt} represents the weight assigned to poll i,
\item
  \(numericgrade_i\) is the numeric grade of poll i, and
\item
  \(n\) is the total number of polls used in the analysis.
\item
  \(pollscore_i\) is the is the pollscore of poll i which reflects the
  estimated bias of the poll (with more negative values indicating less
  bias)
\item
  \(maxPollscore\) is the maximum pollscore across all polls.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  \(\hat{y}_i\)\hspace{0pt} is the predicted percentage of support for
  Kamala Harris from poll i.
\end{itemize}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-akman}{}}%
Akman, Sena. 2023. \emph{What Is Target Population: Definition \&
Examples}. \url{https://forms.app/en/blog/target-population}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-lubridate}{}}%
Grolemund, Garrett, and Hadley Wickham. 2011. {``Dates and Times Made
Easy with {lubridate}.''} \emph{Journal of Statistical Software} 40 (3):
1--25. \url{https://www.jstatsoft.org/v40/i03/}.

@@

@@

>>>>

\begin{document}
\maketitle
\begin{abstract}
This study employs a polls-of-polls approach to predict support for
Kamala Harris and Donald Trump in the 2024 U.S. Presidential Election,
with a focus on key battleground states to forecast the likely winner of
the electoral college. By aggregating multiple polls and applying a
weighted linear regression model with predictors like pollster
reliability, sample size, state, and recency, we estimate a higher
national support for Trump relative to Harris. Our analysis also reveals
a close competition in battleground states, with Trump holding a narrow
lead in Arizona, Georgia, and North Carolina, while Harris leads in
Michigan, Nevada, Pennsylvania, and Wisconsin. These findings underscore
the value of aggregating polls over relying on individual surveys,
offering a more comprehensive and robust forecast of electoral outcomes.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

@@

This paper is organized as follows: In \ref{sec-data}, we clean the
data, explore summary statistics, plot distributions of key variables,
and examine relationships between variables. In \ref{sec-model}, we
discuss our forecasting approach, model selection, justification for the
chosen model, and the mechanism of deriving poll weights based on
pollster reliability, ultimately presenting our predictions. Finally, in
\ref{sec-discussion}, we address the broader implications of our
findings, acknowledge limitations, and suggest directions for future
work.

\hypertarget{sec-data}{%
\section{Data}\label{sec-data}}

@@

The raw data for this project, sourced from FiveThirtyEight,
(FiveThirtyEight 2024) underwent a series of cleaning steps to prepare
it for analysis. Initially, duplicate rows were removed to ensure that
only unique observations remained, facilitated by the \texttt{janitor}
package (Firke 2023). A new binary variable, `national', was created to
indicate whether a poll was conducted at the national or state level.
Missing values in the `state' column were replaced with ``Not
Applicable,'' and numeric grades were evaluated to filter out
low-quality pollsters, keeping only those with a numeric grade above 1.
This cutoff was selected to retain mid to high-level pollsters for more
reliable results. These steps were performed using functions from the
\texttt{tidyverse} package (Wickham et al. 2019). Dates were also
standardized and converted into a proper format for analysis using the
same package. Polls related to Kamala Harris were retained for further
analysis, and percentage support values were transformed into actual
numbers of supporters based on sample size. Additionally, pollster
counts below five were excluded to focus on more reliable data sources.
Polls regarding Kamala Harris were filtered to include only those
conducted after her official candidacy announcement on July 21, 2024,
ensuring the data reflects post-announcement public sentiment.The
cleaned dataset was saved in Parquet format for efficient storage and
retrieval, using the \texttt{arrow} package (Richardson et al. 2024).

@@

\hypertarget{summary-statistics-of-key-variables-and-measurement-method}{%
\subsubsection{Summary Statistics of Key Variables and Measurement
Method}\label{summary-statistics-of-key-variables-and-measurement-method}}

@@

For the purposes this analysis, a subset of variables was selected from
the raw data set. Two additional variables, `national' and `state', were
created using the existing `state' and `end date' variable. A short
description of each variable of interest and their respective
measurement methods is given below.

\begin{itemize}
\item
  \emph{Pollster}: The name of the polling organization conducting the
  poll (e.g., YouGov, RMG Research). This variable helps adjust for
  poll-specific biases. Every pollster that publicly published a
  scientific poll about the 2024 U.S. Presidential Election is included
  in the data set. The value measured for this variable is the name that
  appears in the scientific poll (Morris 2024b).
\item
  \emph{Numeric Grade}: A numeric rating given to the pollster,
  representing the accuracy and methodological transparency of the
  organization (e.g., 3.0), with higher grades indicating more reliable
  pollsters. The value of this variable is calculated by the 538 team.
  All national and state-level polls that were conducted in 1998 or
  later are considered to determine and to assign a numeric grade to
  each pollster (Morris 2024a).
\item
  \emph{Pollscore}: A score that reflects the reliability of each
  pollster, capturing their historical accuracy and biases. Negative
  values indicate better predictive accuracy, and rewards pollsters who
  are accurate and precise. This variable is calculated by aggregating
  predictive error and predictive bias together (Morris 2024a).
  Predictive error is the difference between expectations and reality
  (Renewable Energy 2012) and predictive bias is the systematic bias
  introduced in the methodology (Aguinis and Culpepper 2024).
\item
  \emph{State}: The U.S. state where the poll was conducted, allowing
  for analysis of regional differences in candidate support. The value
  measured for this variable is the primary state that appears in the
  scientific poll's methodology (Morris 2024b).
\item
  \emph{National}: A binary variable indicating whether the poll is
  national (1 for national polls, 0 for state polls).
\item
  \emph{End Date}: The date the poll ended, reflecting the currency of
  the data. This variable is taken from the publication, and is the last
  date for which the survey was active (Morris 2024b).
\item
  \emph{Sample Size}: The total number of respondents in the poll. This
  is measure by the number of completed surveys (Morris 2024b).
\item
  \emph{Candidate Name}: The name of the candidate being polled (e.g.,
  Kamala Harris or Donald Trump), identifying the focus of each poll
  result.
\item
  \emph{Pct (Percentage)}: The percentage of respondents in the poll who
  support the specified candidate.
\item
  \emph{Recency}: This variable measures how recent each poll is. This
  variable is calculated by subtracting the end date of the poll from
  the end date of the most recent poll.
\end{itemize}

@@

Table~\ref{tbl-sumstat} shows that the average poll in our dataset has a
numeric grade of 2.2. This suggests most polls are of moderate to high
quality in terms of reliability. The mean pollscore of -0.5 suggests
these polls are relatively accurate, as negative values imply reduced
bias. Additionally, the large average sample size of 1908 respondents
across polls provides a stable basis for our model, reducing variability
and ensuring reliable predictions for candidate support.

@@

Figure~\ref{fig-ss} shows a clear right-skewed distribution. Most of the
sample sizes are clustered between 0 and 3000 respondents, with a sharp
peak around 1000-1500 respondents. This indicates that the majority of
polls have smaller sample sizes. As sample size increases, the frequency
significantly drops, with very few polls conducted with sample sizes
larger than 5000, though there are a few outliers with sizes approaching
10,000 or more. This wide range in sample sizes can affect the precision
of estimates across different polls.

@@

Figure~\ref{fig-reli} shows the distribution of Numeric Grade (left) and
Pollscore (right) across polling organizations. The Numeric Grade
distribution shows that most pollsters are rated between 1.5 and 3, with
peaks around 2.0 and 2.5, indicating a concentration of pollsters with
moderate to high reliability scores. In contrast, the Pollscore
distribution, where lower values indicate higher reliability, shows a
range primarily between -1.5 and 0, with a notable peak around -0.5.
This suggests that while many polls demonstrate relatively low bias,
there is still variability in reliability across organizations. The
distinction between these two metrics emphasizes the need to consider
both quality (Numeric Grade) and potential systematic bias (Pollscore)
when weighting polls in the model.

@@

Figure~\ref{fig-pie} illustrates the distribution of polls between
candidates and poll types. The left chart shows that the majority of
polls are conducted at the state level, with a smaller portion being
national. The right chart shows that polling is almost evenly split
between Kamala Harris and Donald Trump. This distribution underscores
the model's balanced approach to capturing state-level nuances as well
as broader national trends, providing a comprehensive view of candidate
support across different contexts

@@

\begin{itemize}
\item
  \(y_i\) is the percentage of support for candidate in poll i,
\item
  \(α\) is the intercept,
\item
  \(β_1\) captures the effect of the polling organization,
\item
  \(β_2\) captures the effect of the sample size,
\item
  \(β_3\) captures the effect of recency (how recent the poll is),
\item
  \(β_4\) capture the effects of the different states
\item
  \(\epsilon_i\) represents the error term, assumed to follow a normal
  distribution with mean 0.
\end{itemize}

@@

\begin{itemize}
\item
  \(w_i\) represents the weight assigned to poll i,
\item
  \(numericgrade_i\) is the numeric grade of poll i, and
\item
  \(n\) is the total number of polls used in the analysis.
\item
  \(pollscore_i\) is the is the pollscore of poll i which reflects the
  estimated bias of the poll (with more negative values indicating less
  bias)
\item
  \(maxPollscore\) is the maximum pollscore across all polls.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  \(\hat{y}_i\) is the predicted percentage of support for Kamala Harris
  from poll i.
\end{itemize}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-def_pred_bias}{}}%
Aguinis, Herman, and Steven Culpepper. 2024. \emph{Improving Our
Understanding of Predictive Bias in Testing}.
\url{https://psycnet.apa.org/fulltext/2024-15734-001.html}.

\leavevmode\vadjust pre{\hypertarget{ref-akman}{}}%
Akman, Sena. 2023. \emph{What Is Target Population: Definition \&
Examples}. \url{https://forms.app/en/blog/target-population}.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-538_morris}{}}%
Morris, Elliott. 2024a. \emph{How 538's Pollster Ratings Work}.
\url{https://abcnews.go.com/538/538s-pollster-ratings-work/story?id=105398138}.

\leavevmode\vadjust pre{\hypertarget{ref-538_methodology}{}}%
---------. 2024b. \emph{Trump Leads in Swing-State Polls and Is Tied
with Biden Nationally}.
\url{https://abcnews.go.com/538/trump-leads-swing-state-polls-tied-biden-nationally/story?id=109506070}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-def_pred_error}{}}%
Renewable Energy. 2012. \emph{Prediction Error}.
\url{https://www.sciencedirect.com/topics/engineering/prediction-error}.
